Parsing Postfix log files

### 2008-03-04 doc/logparser.tex
Further corrections after proofreading.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 1000 users.  Coverage is described both in terms of
the fraction of lines parsed and the fraction of mails and connections
successfully reconstructed by the parsing algorithm, including dealing with
false negatives and a discussion of the difficulties in identifying false
positives.  A random sampling of log lines was parsed, and the correctness
of the results manually verified, as part of determining the coverage of
the parser.

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
\sectionref{rule attributes},).  This section continues with a discussion
of rule efficiency concerns, referring to the graphs in
\sectionref{graphs}, and finishes with an explanation of the algorithm used
to generate a new \regex{} from unparsed lines.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules must
        already have been tried and failed on the current line.}  Any given
        rule will only be used to parse a subset of the log lines, and any
        given log line will only be parsed by a subset of the rules.

    \item The first matching rule wins: no further rules are tried against
        that line, but there is a facility for specifying the order of
        rules so that more specific rules can be tried first.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{}, and determine if any overlap.  The
author has not attempted to do this: such a project by itself would
probably qualify for a PhD, and may involve solving the Halting
Problem~\cite{Wikipedia-halting-problem} and circumventing the
Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line is unspecified, and
the first matching rule wins.  Overlapping rules are frequently a
requirement, allowing a more specific rule to match some lines and a more
general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

Detecting overlapping rules is difficult, but the following may be helpful:

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See \sectionref{rule efficiency} for more
        details of sorting the rules.}  Parse a number of log files using
        normal sorting, dump a textual representation of the connections
        and results tables, and delete everything from those tables.
        Repeat with shuffled and reversed sorting.  If there are no
        overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules.  Which rules overlap can be
        determined by examining the differences in the tables: each result
        contains the id of the rule which created it, so the rule
        referenced in the normal table overlaps with the rule referenced in
        the reversed table.  Unfortunately this method cannot prove the
        absence of overlapping rules; it can detect overlapping rules, but
        only if there are log lines in the input files which match more
        than one rule.

@@

The various fields are used as follows;

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the \regex{} will be tried against
        the line; if the match is successful the action will be executed,
        if not the next rule will be tried.  If the program-specific rules
        don't match the log line, the generic rules will be tried in the
        same way.

@@

    \item [queueid:] The index of the field in the \regex{} which captured
        the queueid, or zero if the line does not contain a queueid.  This
        allows the correct mail can be found by queueid and actions
        performed on it.

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see \sectionref{rule ordering for efficiency});
        hits is set to the number of successful matches at the end of the
        parsing run, and hits\_total has hits added to it, but is otherwise
        unused by the algorithm.

@@

Additional fields are captured by \_\_RESTRICTION\_START\_\_, hence the
fields in result\_cols and connection\_cols start at 5 in the example.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (for a small mail
server --- large scale mail servers would have much larger log files on a
daily basis).  When generating the data for the graphs included in
appendix~\refwithpage{graphs}, \numberOFlogFILES{} log files (totaling
10.08 GB, 60.72 million lines) were each parsed 10 times, the first run
discarded, and the remaining 9 runs averaged.  The first run is discarded
for two reasons:

@@

Saving results to the database was disabled for the test runs, as that
dominates the run time of the program, and the tests are aimed at measuring
the speed of the parser rather than the speed of the database.

@@

An important property of a parser is how execution time scales with input
size: does it scale linearly, polynomially, or exponentially?
Graph~\refwithpage{execution time vs file size vs number of lines graph}
shows the execution time in seconds, file size in MB and tens of thousands
of lines per log file per log file.  All three lines run roughly in
parallel, giving a visual impression that the algorithm scales linearly
with input size.  This impression is borne out by
graph~\refwithpage{execution time vs file size vs number lines factor},
which plots the ratio of file size vs execution time and number of lines vs
execution time.  As the reader can see, the ratios are quite tightly
banded, showing that the algorithm scales linearly: the much larger log
files between points 60 and 70 on the X axis in
graph~\refwithpage{execution time vs file size vs number of lines graph}
don't cause any abnormality in the corresponding points in
graph~\refwithpage{execution time vs file size vs number lines factor}.

@@

    \item [reverse] Hypothetically the least optimal order: the most
        frequently matching rules will be tried last.

@@

Graphs~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal} show the
percentage increase of execution times (mean and standard deviation are
shown in table~\refwithpage{Regex caching/discarding with different groups
of log files}).  Overall this provides a modest but worthwhile
performance increase.

@@

The core of the algorithm is quite simple: log lines are generally created
by substituting variable words into a fixed pattern, and analysis of the
frequency with which each word occurs can be used to determine whether the
word is fixed or variable.  This classification can be used to group
similar log lines and generate a \regex{} to match each group of lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this phase is to utilise the existing knowledge to create more
        accurate \regexes{}.  The new log lines are written to a temporary
        file, which all subsequent stages use instead of the original input
        file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each line, incrementing a counter for each hash.  The counters
        will later be used to filter out words: if the word's hash does not
        have a high frequency, the word itself cannot have a high
        frequency, and there is no need to maintain a counter for it,
        reducing the number of counters and thus the programme's memory
        consumption.

    \item [Calculate word frequencies]  The position of words within a line
        is important: a common word does not indicate similarity between
        lines unless it occupies the same position within both
        lines.\footnote{If a variable term within a line contains spaces,
        it will appear to the algorithm as two words rather than one.  This
        will alter the position of subsequent words, so a word occurring in
        different positions in two log lines may indicate similarity, but
        the algorithm does not attempt to deal with this possibility.}  The
        algorithm maintains a counter for each \textit{(word, word's
        position within the line)\/} tuple, incrementing it each time that
        word occurs in that position.

        The original algorithm only maintains counters for words whose hash
        result has a high frequency from the previous phase; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm.  The modified algorithm omits
        this check because the majority of unique or infrequently occurring
        words will have been substituted with keywords during the first
        phase, vastly reducing the number of tuples to maintain counters
        for.

    \item [Classify words based on their frequency]  The frequency of each
        \textit{(word, word's position within the line)\/} tuple is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all lines is generally a good starting point) it
        is classified as a fixed word, otherwise it is classified as a
        variable term.  Variable terms are replaced by \texttt{.+}, which
        means to match zero or more of any character.

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the line, and a counter is maintained for each \regex{}.
        Contiguous sequences of \texttt{.+} in the newly reassembled
        \regexes{} are collapsed to a single \texttt{.+}; any resulting
        duplicate \regexes{} are combined, and their counters summed.  If
        the frequency of a \regex{} is lower than the threshold supplied by
        the user the \regex{} is discarded.\footnote{This is a second,
        independent threshold, but once again 1\% of input lines is a good
        starting point; obviously the threshold depends on the number and
        type of input lines.}  The new \regexes{} are printed for the user
        to add to the database, either as new rules or merged into the
        \regexes{} of existing rules; the counter for each \regex{} is also
        printed, giving the user an indication of how many of the input
        lines that \regex{} should match.  Discarding \regexes{} will
        result in some of the input lines not being matched; this utility
        should be run again once the unmatched lines have been isolated by
        running the parser, including the new \regexes{}, with the same
        input which produced the original set of unparsed lines.

@@

    \item An example log line and the rule matching it illustrated a
        description of how the fields in the rule are used both in the
        matching phase and the subsequent action that is executed.

    \item The database table containing the rules is dealt with in
        \sectionref{rule attributes}, and is not duplicated in this
        section.

@@

@@

The difference between logs with a mail loop and logs without can be seen
in table~\refwithpage{Rule ordering with and without a mail loop} showing
the increases for the different rule orderings and combinations of log
files:

\showtable{build/stats-normal-shuffle-reverse-include}{Rule ordering with
and without a mail loop}{Rule ordering with and without a mail loop}

>>>>

XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 700 users.  Coverage is described both in terms of
the fraction of lines parsed and the fraction of mails and connections
successfully reconstructed by the parsing algorithm, including dealing with
false negatives and a discussion of the difficulties in identifying false
positives.  A random sampling of log lines was parsed, and the correctness
of the results manually verified, as part of determining the coverage of
the parser.

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, followed by the problems caused by
overlapping rules, and techniques to detect and deal with such rules.  An
example rule and a log line it would match are provided, plus a description
of how the fields in the rule\footnote{The details of the table containing
the rules have already been described in \sectionref{rule attributes}.} are
used when matching a line and subsequently performing the requested action.
The section continues with a discussion of rule efficiency concerns,
referring to the graphs in \sectionref{graphs}, and finishes with an
explanation of the algorithm used to generate new \regexes{} from unparsed
lines.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules have
        already have been tried and failed on the current line.}  Any given
        rule will only be used to parse a subset of the log lines, and any
        given log line will only be parsed by a subset of the rules.

    \item The first matching rule wins: no further rules are tried against
        that line, but there is a facility for ordering the rules so that
        more specific rules can be tried first.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{} and determine if any overlap.  The
author has not attempted to do this: such a project by itself would
probably qualify for a PhD, and may involve solving the Halting
Problem~\cite{Wikipedia-halting-problem} and circumventing the
Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line may change between log
files, and the first matching rule wins.  Overlapping rules are frequently
a requirement, allowing a more specific rule to match some lines and a more
general rule to match the majority, e.g.\ separating \SMTP{} delivery to
specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

Detecting overlapping rules is difficult, but the following approaches may
be helpful:

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See \sectionref{rule efficiency} for more
        details of sorting the rules.}  Parse a number of log files using
        normal sorting, then dump a textual representation of the rules,
        connections and results tables.  Repeat with shuffled and reversed
        rules, starting with a fresh database.  If there are no overlapping
        rules the tables from each run will be identical; differences
        indicate overlapping rules.  Which rules overlap can be determined
        by examining the differences in the tables: each result contains a
        reference to the rule which created it, if the references differ
        between runs the two rules referenced in the differing records
        overlap.  Unfortunately this method cannot prove the absence of
        overlapping rules; it can detect overlapping rules, but only if
        there are log lines in the input files which match more than one
        rule.

@@

Additional data will be captured automatically when the \regex{} contains 
\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
connection\_cols start at 5.  The various fields are used as follows;

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line a match using the \regex{} will be
        attempted against the line; if the match is successful the action
        will be executed, if not the next rule will be tried.  If the
        program-specific rules don't match the log line, the parser will
        fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

@@

    \item [queueid:] The index of the capture in the \regex{} which
        supplies the queueid, or zero if the line does not contain a
        queueid.  This allows the correct mail can be found by queueid and
        actions performed on it.

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see \sectionref{rule ordering for efficiency});
        hits is set to the number of successful matches at the end of the
        parsing run; hits\_total is the sum of hits over every parsing run,
        but is otherwise unused by the algorithm.

@@

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (generated daily on
a mail server handling mail for approximately 700 users --- large scale
mail servers would have much larger log files on a daily basis).  When
generating the data for the graphs included in \sectionref{graphs},
\numberOFlogFILES{} log files (totaling 10.08 GB, 60.72 million lines) were
each parsed 10 times, the first run discarded, and the execution time for
the remaining 9 runs averaged.  The first run is discarded for two reasons:

@@

Saving results to the database was disabled for the test runs, as that
dominates the run time of the program, and the tests are aimed at measuring
the speed of the parser rather than the speed of the database and the disks
the database is stored on.

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
Graph~\refwithpage{execution time vs file size vs number of lines graph}
shows the execution time in seconds, file size in MB and tens of thousands
of lines per log file.  All three lines run roughly in parallel, giving a
visual impression that the algorithm scales linearly with input size.  This
impression is borne out by graph~\refwithpage{execution time vs file size
vs number lines factor}, which plots the ratio of file size vs execution
time and ratio of number of lines vs execution time (higher is
better).\footnote{Table~\refwithpage{execution time vs file size vs number
lines factor table} shows the ratios for different breakdowns of the log
files.}  As the reader can see the ratios are quite tightly banded, showing
that the algorithm scales linearly: the much larger log files between
points 60 and 70 on the X axis in graph~\refwithpage{execution time vs file
size vs number of lines graph} actually cause the ratio to increase, rather
than decrease.  The strange behaviour where larger log files are parsed
more efficiently is explained fully in \sectionref{Why are there dips in
the graphs?}.

@@

    \item [reverse] Hypothetically the worst order: the most frequently
        matching rules will be tried last.

@@

Graphs~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal} show the
percentage increase of execution times, with table~\refwithpage{Execution
time increase for different rule orderings} showing the mean increases for
different groupings of logs.  Overall this provides a modest but worthwhile
performance increase of approximately 9\%, for a small investment in time
and programming.

@@

The core of the algorithm is quite simple: log lines are generally created
by substituting variable words into a fixed pattern, and analysis of the
frequency with which each word occurs can be used to determine whether the
word is variable or part of the fixed pattern.  This classification can be
used to group similar log lines and generate a \regex{} to match each group
of lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this phase is to utilise existing knowledge to create more accurate
        \regexes{}.  The new log lines are written to a temporary file,
        which all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the preprocessing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash from a small range of values for each word
        of each line, incrementing a counter for each hash.  The counters
        will be used in the next pass to filter out words: if the word's
        hash does not have a high frequency, the word itself cannot have a
        high frequency, and there is no need to maintain a counter for it,
        reducing the number of counters and thus the program's memory
        consumption.

    \item [Calculate word frequencies]  The position of words within a line
        is important: a common word does not indicate similarity between
        lines unless it occupies the same position within both
        lines.\footnote{If a variable term within a line contains spaces,
        it will appear to the algorithm as two words rather than one.  This
        will alter the position of subsequent words, so a word occurring in
        different positions in two log lines \textit{may\/} indicate
        similarity, but the algorithm does not attempt to deal with this
        possibility.}  The algorithm maintains a counter for each
        \textit{(word, word's position within the line)\/} tuple,
        incrementing it each time that word occurs in that position.

        The original algorithm only maintains counters for words whose hash
        result from the previous phase has a high frequency; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm.  The modified algorithm omits
        this check because the majority of unique or infrequently occurring
        words will have been substituted with keywords during the first
        phase, vastly reducing the number of tuples to maintain counters
        for.

    \item [Classify words based on their frequency]  The frequency of each
        \textit{(word, word's position within the line)\/} tuple is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all lines is generally a good starting point) it
        is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be classified as a fixed term, but that should be noticed by
        the user when reviewing the new \regexes{}.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the line, and a counter is maintained for each \regex{}.
        Contiguous sequences of \texttt{.+} in the newly reassembled
        \regexes{} are collapsed to a single \texttt{.+}; any resulting
        duplicate \regexes{} are combined, and their counters added.  If
        the frequency of a \regex{} is lower than the threshold supplied by
        the user the \regex{} is discarded.\footnote{This is a second
        threshold, independent of the threshold used to differentiate
        between fixed and variable words, but once again 1\% of input lines
        is a good starting point; obviously the threshold depends on the
        number and type of input lines.}  The new \regexes{} are printed
        for the user to add to the database, either as new rules or merged
        into the \regexes{} of existing rules; the counter for each
        \regex{} is also printed, giving the user an indication of how many
        of the input lines that \regex{} should match.  Discarding
        \regexes{} will result in some of the input lines not being
        matched; this utility should be run again once the unmatched lines
        have been isolated by running the parser, including the new
        \regexes{}, with the same input which produced the original set of
        unparsed lines.

@@

    \item An example log line and the rule matching it illustrated a
        description of how the fields in the rule are used both in the
        matching phase and the action that is subsequently executed.

    \item The structure of the database table containing the rules is dealt
        with in \sectionref{rule attributes}, and is not duplicated in this
        section.

@@

\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
increase for different rule orderings}{Execution time increase for
different rule orderings}

@@

The difference between logs with a mail loop and logs without can be seen
in table~\refwithpage{Execution time increase for different rule orderings}
showing the increases for the different rule orderings and combinations of
log files:
